Databases are an effective and efficient method for storage and access of mass-spectrometry data

### 2025-03-11 manuscript/manuscript.Rmd
First draft done and circulated to the lab

method_names <- c("pyteomics", "pymzml", "pyopenms", "pyopenms_2DPeak",
                  "mzMLb", "mzDB", "MZA", "mz5", 
                  "MZTree", "mzMD", 
                  "SQLite", "DuckDB", "Parquet")
method_colors <- c("#0B4F5B", "#147C8F", "#21A8C0", "#44C3DA", 
                   "#028e34", "#f4bb23", "#e0670b", "#a41118", 
                   "#E25C76", "#E891A2", 
                   "#580770", "#8A1BAC", "#B04ACF")

@@

## Mass-spectrometry files

@@

Files were downloaded as .raw. mzML, mz5, and mzMLb were all natively supported by Proteowizard's `msconvert` software (version 3.0.25009) while MZA (v1.24.11.16) and mzDB (v0.9.10_build20170802) had separate extensions to this executable enabling their own conversion. MzTree and mzMD were converted via their GUI which did not have release or versioning information available but were downloaded from Github (https://github.com/optimusmoose/MZTree and https://github.com/yrm9837/mzMD-java, respectively) and built via Maven (v3.9.9) for Java (v21.0.6). SQL databases were built using Python 3.11.11 with SQLite (v3.48.0) via the Python sqlite3 package (v2.6.0), DuckDB via the duckdb package (v1.1.3), and Parquet files via the pyarrow package (v19.0.0). We were additionally able to download and install []

@@

```{shell, eval=FALSE}
cd Documents\Will
git clone https://github.com/wkumler/mzsql
cd mzsql

@@

The “vanilla” database style proposed here abandons a 1:1 representation of the original vendor-specific file. This decision was made after discussion with a wide variety of experts, all of whom preserved the original MS files even after conversion to another file type, indicating that a highly-performant addition is more important than direct replacement. Here, we map MS concepts (retention time, *m/z*, intensity, etc.) directly to database fields to make downstream processing as intuitive as possible. Metadata is stored separately in `file_info` and `scan_info` tables that are linked by filename and scan number (Figure XX). We do not force compression of any of these fields because we believe that decoding compressed data is both a slow and unintuitive step, though automatic compression is supplied by the DuckDB and Parquet file types.

```{r gtables for figure 1}
library(gt)
data.frame(
  filename=c("Smp_A", "Smp_B", "Smp_Y", "Smp_Z"),
  n_scans=c(1445, 1489, 1388, 1390),
  timestamp=c("2021-08-23 04:30", "2021-08-23 05:25", "2021-08-23 22:15", "2021-08-23 23:10"),
  `etc.`=c("...", "...", "...", "...")
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             n_scans=html("n_scans</br><small>Scan count</small>"),
             timestamp=html("timestamp</br><small>Timestamp</small>"),
             `etc.`=html("Etc.</br><small>...</small>")) %>%
  tab_row_group(label = "...additional samples...", rows = !c(TRUE, TRUE, FALSE, FALSE)) %>%
  tab_row_group(label = "", rows = c(TRUE, TRUE, FALSE, FALSE)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: file_info") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/fileinfo_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(1, 2, 1389, 1390),
  ms_level=c(1, 2, 2, 2),
  rt=c(0.10, 0.12, 22.43, 22.45),
  # min_mz=c(59.99975, 49.99992, 50.00013, 49.99978),
  TIC=c(2331809, 820427, 9555703, 1892338),
  `etc.`=c("...", "...", "...", "...")
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             ms_level=html("ms_level</br><small>MS level</small>"),
             rt=html("rt</br><small>Retention time</small>"), 
             # min_mz=html("min_mz</br><small>Lowest <emph>m/z</emph></small>"),
             TIC=html("TIC</br><small>Ion current</small>"),
             `etc.`=html("Etc.</br><small>...</small>")) %>%
  tab_row_group(label = "...many additional scans...", rows = !c(TRUE, TRUE, FALSE, FALSE)) %>%
  tab_row_group(label = "", rows = c(TRUE, TRUE, FALSE, FALSE)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: scan_info") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/scaninfo_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(1,1,1385,1385),
  rt=c(0.10, 0.10, 22.35, 22.35),
  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
  int=c(6618, 2657, 158084, 4673)
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             rt=html("rt</br><small>Retention time</small>"), 
             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"), 
             int=html("int</br><small>Intensity</small>")) %>%
  tab_row_group(label = "...millions of additional entries...", rows = rt>=mean(rt)) %>%
  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: MS1") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/ms1_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(2,2,1390,1390),
  prescan=c(1,1,1385,1385),
  ms2_rt=c(0.12, 0.12, 22.45, 22.45),
  fragmz=c(51.0238, 53.0394, 52.0186, 57.0923),
  premz=c(241.0894, 241.0894, 185.1932, 185.1932),
  ms2_int=c(36104, 243165, 28371, 129604)
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             prescan=html("prescan</br><small>Precursor scan</small>"),
             ms2_rt=html("rt</br><small>Retention time</small>"), 
             fragmz=html("fragmz</br><small>Fragment <emph>m/z</emph></small>"), 
             premz=html("premz</br><small>Precursor <emph>m/z</emph></small>"), 
             ms2_int=html("int</br><small>Intensity</small>")) %>%
  tab_row_group(label = "...millions of additional entries...", rows = ms2_rt>=mean(ms2_rt)) %>%
  tab_row_group(label = "", rows = ms2_rt<=mean(ms2_rt)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: MS2") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "figures/ms2_table.png", vwidth=1000)
```

@@

*Figure XX: Database schema for an example MS/MS dataset showing the organization of mass-spectrometry data into tables. Fields of interest are easily queryable with simple SQL commands as shown in the table at bottom.*

@@

[TODO: Note the method used to identify the m/zs, scans, and RT ranges extracted here]

Via %timeit% (need to redo these with randomization)

@@

```{r singlefile_fig}
all_timings <- read_csv("data/singlefile_times.csv") %>% 
  mutate(method=factor(method, levels=method_names)) %>%
  mutate(query=factor(query, levels=c("ms1_spec", "chrom", "rtrange",
                                      "ms2_spec", "premz", "fragmz"),
                      labels=c("Full scan", "Chromatogram", "RT Range",
                               "MS/MS scan", "Precursor search",
                               "Fragment search")))
all_sizes <- read_csv("data/file_sizes.csv") %>%
  mutate(method=factor(method, levels=method_names))

time_boxplot <- all_timings %>%
  ggplot() +
  geom_boxplot(aes(x=method, y=time, color=method)) +
  facet_wrap(~query) +
  scale_y_log10() +
  scale_color_manual(breaks = method_names, values=method_colors) +
  labs(x=NULL, y="Time (seconds)", color=NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
        legend.position = "none")
size_barplot <- all_sizes %>%
  filter(!method%in%c("pyteomics", "pyopenms", "pyopenms_2DPeak")) %>%
  mutate(method=as.character(method)) %>%
  mutate(method=ifelse(method=="pymzml", "mzML", method)) %>%
  mutate(method=factor(method, levels=c("mzML", method_names))) %>%
  mutate(query="File size") %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=method, y=file_size/1e6, fill=method), color="black") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(breaks = c("mzML", method_names), 
                    values=c("#147C8F", method_colors)) +
  facet_wrap(~query) +
  labs(x=NULL, y="File size (MB)", fill="File type") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5), legend.position = "none")

@@

ggsave("figures/singlefile_fig.png", width = 6.5, height=5)

@@

*Figure XX: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper. The left six panels show boxplots representing the time required in seconds to extract an full scan spectrum, an ion chromatogram, all data within a retention time range, an MS/MS scan, the fragments of a specified precursor, and all precursors with a specified fragment. The error in the boxplot is composed of timing information for 10 repeated queries, each of a different target scan number, retention time (RT), or m/z. The right panel shows a barplot of the size on disk in megabytes (MB) occupied by each file type.*

```{r singlefile_supp}
fun_levels = c("<span style='color:#a41118'>Including initial load</span>", 
               "<span style='color:#0b505c'>Extraction only</span>")
init_outside_gp <- read_csv("data/singlefile_supp.csv") %>%
  pivot_longer(pyteomics:mz5, names_to = "method", values_to="time") %>%
  mutate(method=factor(method, levels=method_names)) %>%
  mutate(fun_type=factor(fun_type, levels=c("inclusive", "minimal"),
                         labels=fun_levels)) %>%
  ggplot() +
  geom_boxplot(aes(x=method, y=time, color=fun_type)) +
  scale_y_log10() +
  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
  labs(x=NULL, y="Time (seconds)", color=NULL) +
  theme_bw() +
  theme(legend.text=element_markdown(), legend.position = "inside",
        legend.justification.inside = c(0.98, 0.95),
        legend.background = element_rect(color='black'),
        plot.title.position = "plot", 
        plot.caption.position =  "plot") +
  ggtitle("A) Scan extraction with and without pre-loading")

pyopenms_precomp_gp <- read_csv("data/pyopenms_precomp.csv") %>%
  pivot_longer(rep1:rep3) %>%
  mutate(type=factor(type, levels=c("init", "precomp"),
                       labels=fun_levels)) %>%
  mutate(method=factor(method, levels=c("spec", "chrom", "chrom_2D", "rtrange",
                                        "rtrange_2D", "premz", "fragmz"),
                       labels = c("Spectrum", "EIC", "EIC\n(2D Peak method)",
                                  "RT Range", "RT Range\n(2D Peak method)",
                                  "Precursor\nsearch", "Fragment\nsearch"))) %>%
  ggplot() +
  geom_boxplot(aes(x=method, y=value, color=type)) +
  scale_y_log10() +
  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
  theme_bw() +
  theme(legend.position = "none", 
        plot.title.position = "plot", 
        plot.caption.position =  "plot") +
  labs(x=NULL, y="Time (seconds)", color=NULL) +
  ggtitle("B) pyopenms metric timings with and without pre-loading")

init_outside_gp / pyopenms_precomp_gp
ggsave("figures/supp_preload_fig.png", width = 6.5, height=5)
```

@@

We also investigated the efficacy of the various MS data formats for MS/MS data and found that support for fragmentation data searches was lacking or absent from the documentation and exposed functionality of each of these file types, requiring custom implementations every time. Despite both precursor searches (where all the precursors of a given fragment are found) and fragment searches (where all the fragments of a given precursor are found) representing intuitive and useful methods of MS/MS data processing, these timings were consistently among the slowest of the six query types for the non-database methods (Figure XX, rightmost panels).

All existing MS data types required multiple seconds to perform a single fragment search (Figure XX, bottom-right panel), representing a significant bottleneck for any downstream analysis requiring the data associated with the fragments of a given precursor. The SQL-based parsers, on the other hand, all took fractions of a second and consistently returned the relevant data hundreds of times more quickly than existing methods. The same was true for a precursor search across all methods aside from mzDB (Figure XX, bottom middle panel), which benefited significantly from constructing a single bounding box for all MS/MS information that requires a single decoding into computer memory, though this strategy will fail for any file with sufficiently large MS/MS data.

@@

File size is another important constraint on the efficacy of various MS formats. We measured the size on disk of each of the file types and found that they varied by approximately an order of magnitude, with HDF-based file types hovering around one-third the size of the mzML (83 megabytes, MB) while mzDB and mzMD were larger (108 MB and 96 MB, respectively). The SQLite object was the largest on disk of all the file types, nearly doubling the mzML's size at 145 MB, while DuckDB was on par with MZA at two-thirds of the mzML (54 MB) and Parquet improved slightly upon that again with its columnar-based storage format (Figure XX).

@@

The single-file, single-metric case discussed above and shown in Figure XX is largely a worst-case scenario for many MS data systems that have a slow initial setup step to make downstream analysis faster. To compare these systems more fairly to our database schema, we also tested timings across multiple chromatograms. In each case, this was implemented as a for loop iterating over an increasing number of chromatograms corresponding to the largest intensity ions in the file (Figure XX).

```{r}
multichrom_times <- read_csv("data/multichrom_times.csv")
timelabs <- c("0.01 seconds", "One second", "One minute", "One hour")
multichrom_times %>%
  mutate(method=factor(method, levels=method_names)) %>%
  ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
  geom_abline(slope=1, intercept=-3:3, color="black") +
  geom_smooth(method="lm", formula="y~x", alpha=0.2, fill="grey50") +
  geom_point(position = position_dodge(width=0.1), color="black", pch=21, size=3) +
  scale_y_log10(breaks=c(0.01, 1, 60, 3600), labels=timelabs) +
  scale_x_log10(breaks=c(1, 3, 10, 30, 100)) +
  scale_fill_manual(breaks = method_names, values=method_colors, 
                    aesthetics = c("fill", "color")) +
  labs(x="Number of chromatograms extracted", y="Time required",
       color=NULL, fill=NULL) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
ggsave("figures/multichrom_fig.png", width = 6.5, height=5)
```

@@

*Figure XX: Scatter plot of the time required to extract multiple chromatograms using various methods on logarithmic axes. Best-fit linear models have been added for each method are shown behind triplicate timing measurements. Transparent intervals around each best-fit line show a single standard error of the mean. Chromatograms correspond to the largest intensity ions in the file. 1:1 lines have been added in black behind the data for comparison.*

Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure XX). pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30. Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure XX).

>>>>

method_names <- c("pyteomics", "pymzml", "pyopenms", "pyopenms_2DPeak",
                  "mzMLb", "mzDB", "MZA", "mz5", 
                  "MZTree", "mzMD", 
                  "SQLite", "DuckDB", "Parquet")
method_colors <- c("#393B79", "#5254A3", "#6B6ECF", "#9C9EDE", 
                   "#0F8299", "#637939", "#E6550D", "#843C39", 
                   "#7B4173", "#CE6DBD", 
                   "#8C6D31", "#E7BA52", "#efddb9")

@@

## Mass-spectrometry files and software used

@@

Files were downloaded as .raw. mzML, mz5, and mzMLb were all natively supported by Proteowizard's `msconvert` software (version 3.0.25009) while MZA (v1.24.11.16) and mzDB (v0.9.10_build20170802) had separate extensions to this executable enabling their own conversion. MzTree and mzMD were converted via their GUI which did not have release or versioning information available but were downloaded from Github (https://github.com/optimusmoose/MZTree and https://github.com/yrm9837/mzMD-java, respectively) and built via Maven (v3.9.9) for Java (v21.0.6). SQL databases were built using Python 3.11.11 with SQLite (v3.48.0) via the Python sqlite3 package (v2.6.0), DuckDB via the duckdb package (v1.1.3), and Parquet files via the pyarrow package (v19.0.0). We were additionally able to download and install [blah]

@@

```{r shell, eval=FALSE}
cd Documents\Will
git clone https://github.com/wkumler/mzsql
cd mzsql

@@

The “vanilla” database style proposed here abandons a 1:1 representation of the original vendor-specific file. This decision was made after discussion with a wide variety of experts, all of whom preserved the original MS files even after conversion to another file type, indicating that a highly-performant addition is more important than direct replacement. Here, we map MS concepts (retention time, *m/z*, intensity, etc.) directly to database fields to make downstream processing as intuitive as possible. Metadata is stored separately in `file_info` and `scan_info` tables that are linked by filename and scan number (Figure 1). We do not force compression of any of these fields because we believe that decoding compressed data is both a slow and unintuitive step, though automatic compression is supplied by the DuckDB and Parquet file types.

```{r gtables for figure 1, eval=FALSE}
library(gt)
data.frame(
  filename=c("Smp_A", "Smp_B", "Smp_Y", "Smp_Z"),
  n_scans=c(1445, 1489, 1388, 1390),
  timestamp=c("2021-08-23 04:30", "2021-08-23 05:25", "2021-08-23 22:15", "2021-08-23 23:10"),
  `etc.`=c("...", "...", "...", "...")
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             n_scans=html("n_scans</br><small>Scan count</small>"),
             timestamp=html("timestamp</br><small>Timestamp</small>"),
             `etc.`=html("Etc.</br><small>...</small>")) %>%
  tab_row_group(label = "...additional samples...", rows = !c(TRUE, TRUE, FALSE, FALSE)) %>%
  tab_row_group(label = "", rows = c(TRUE, TRUE, FALSE, FALSE)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: file_info") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "manuscript/figures/fileinfo_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(1, 2, 1389, 1390),
  ms_level=c(1, 2, 2, 2),
  rt=c(0.10, 0.12, 22.43, 22.45),
  # min_mz=c(59.99975, 49.99992, 50.00013, 49.99978),
  TIC=c(2331809, 820427, 9555703, 1892338),
  `etc.`=c("...", "...", "...", "...")
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             ms_level=html("ms_level</br><small>MS level</small>"),
             rt=html("rt</br><small>Retention time</small>"), 
             # min_mz=html("min_mz</br><small>Lowest <emph>m/z</emph></small>"),
             TIC=html("TIC</br><small>Ion current</small>"),
             `etc.`=html("Etc.</br><small>...</small>")) %>%
  tab_row_group(label = "...many additional scans...", rows = !c(TRUE, TRUE, FALSE, FALSE)) %>%
  tab_row_group(label = "", rows = c(TRUE, TRUE, FALSE, FALSE)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: scan_info") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "manuscript/figures/scaninfo_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(1,1,1385,1385),
  rt=c(0.10, 0.10, 22.35, 22.35),
  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
  int=c(6618, 2657, 158084, 4673)
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             rt=html("rt</br><small>Retention time</small>"), 
             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"), 
             int=html("int</br><small>Intensity</small>")) %>%
  tab_row_group(label = "...millions of additional entries...", rows = rt>=mean(rt)) %>%
  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: MS1") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "manuscript/figures/ms1_table.png", vwidth=1000)

data.frame(
  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
  scan_num=c(2,2,1390,1390),
  prescan=c(1,1,1385,1385),
  ms2_rt=c(0.12, 0.12, 22.45, 22.45),
  fragmz=c(51.0238, 53.0394, 52.0186, 57.0923),
  premz=c(241.0894, 241.0894, 185.1932, 185.1932),
  ms2_int=c(36104, 243165, 28371, 129604)
) %>%
  gt() %>%
  cols_label(filename=html("filename</br><small>Source file</small>"),
             scan_num=html("scan_num</br><small>Scan number</small>"),
             prescan=html("prescan</br><small>Precursor scan</small>"),
             ms2_rt=html("rt</br><small>Retention time</small>"), 
             fragmz=html("fragmz</br><small>Fragment <emph>m/z</emph></small>"), 
             premz=html("premz</br><small>Precursor <emph>m/z</emph></small>"), 
             ms2_int=html("int</br><small>Intensity</small>")) %>%
  tab_row_group(label = "...millions of additional entries...", rows = ms2_rt>=mean(ms2_rt)) %>%
  tab_row_group(label = "", rows = ms2_rt<=mean(ms2_rt)) %>%
  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
  opt_table_outline() %>%
  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
  cols_align(align = "center") %>%
  cols_move_to_start("filename") %>%
  tab_options(row_group.padding = 3) %>%
  tab_header(title = "Table: MS2") %>%
  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
  gtsave(filename = "manuscript/figures/ms2_table.png", vwidth=1000)
```

@@

*Figure 1: Database schema for an example MS/MS dataset showing the organization of mass-spectrometry data into tables. Fields of interest are easily queryable with simple SQL commands as shown in the table at bottom.*

*Supplemental figure XX: (TBD) Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imagining.*

@@

We randomly sampled a single file out of the full dataset for comparison across metrics and file formats (20220923_LEAP-POS_QC04). We sampled 100 random scan numbers using SQLite's ORDER BY RANDOM() function and pulled out the largest ions for chromatogram extraction using a 10 ppm mass range window. Retention time ranges were the highest-intensity retention time of each ion chromatogram plus or minus one minute. Similarly, the largest fragments by intensity were used for the MS/MS metrics with a 10 ppm mass range exclusion window.

Timing was performed via Python's timeit library and the timeit.repeat function, with the various file formats as the innermost loop to ensure bias over time was distributed equally among function calls. File sizes were estimated using Python's os library with os.path.getsize. We did not exhaustively monitor memory usage, though our heuristic exploration of the timing scripts did not ever indicate that memory was a constraint.

Timing data was obtained on a Lenovo XX blah XX with [cores] and [memory].

@@

```{r singlefile_fig}
all_timings <- read_csv("manuscript/data/singlefile_times.csv") %>% 
  mutate(method=factor(method, levels=method_names)) %>%
  mutate(query=factor(query, levels=c("ms1_spec", "chrom", "rtrange",
                                      "ms2_spec", "premz", "fragmz"),
                      labels=c("Full scan", "Chromatogram", "RT Range",
                               "MS/MS scan", "Precursor search",
                               "Fragment search")))
all_sizes <- read_csv("manuscript/data/file_sizes.csv") %>%
  mutate(method=factor(method, levels=method_names))

time_boxplot <- all_timings %>%
  ggplot() +
  geom_boxplot(aes(x=method, y=time, color=method)) +
  facet_wrap(~query) +
  scale_y_log10() +
  scale_color_manual(breaks = method_names, values=method_colors) +
  labs(x=NULL, y="Time (seconds)", color=NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
        legend.position = "none")
size_barplot <- all_sizes %>%
  filter(!method%in%c("pyteomics", "pyopenms", "pyopenms_2DPeak")) %>%
  mutate(method=as.character(method)) %>%
  mutate(method=ifelse(method=="pymzml", "mzML", method)) %>%
  mutate(method=factor(method, levels=c("mzML", method_names))) %>%
  mutate(query="File size") %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=method, y=file_size/1e6, fill=method), color="black") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(breaks = c("mzML", method_names), 
                    values=c("#393B79", method_colors)) +
  facet_wrap(~query) +
  labs(x=NULL, y="File size (MB)", fill="File type") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5), legend.position = "none")

@@

ggsave("manuscript/figures/singlefile_fig.png", width = 6.5, height=5)

@@

*Figure 2: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper. The left six panels show boxplots representing the time required in seconds to extract an full scan spectrum, an ion chromatogram, all data within a retention time range, an MS/MS scan, the fragments of a specified precursor, and all precursors with a specified fragment. The error in the boxplot is composed of timing information for 10 repeated queries, each of a different target scan number, retention time (RT), or m/z. The right panel shows a barplot of the size on disk in megabytes (MB) occupied by each file type.*

```{r singlefile_supp}
fun_levels = c("<span style='color:#a41118'>Including initial load</span>", 
               "<span style='color:#0b505c'>Extraction only</span>")
init_outside_gp <- read_csv("manuscript/data/singlefile_supp.csv") %>%
  pivot_longer(pyteomics:mz5, names_to = "method", values_to="time") %>%
  mutate(method=factor(method, levels=method_names)) %>%
  mutate(fun_type=factor(fun_type, levels=c("inclusive", "minimal"),
                         labels=fun_levels)) %>%
  ggplot() +
  geom_boxplot(aes(x=method, y=time, color=fun_type)) +
  scale_y_log10() +
  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
  labs(x=NULL, y="Time (seconds)", color=NULL) +
  theme_bw() +
  theme(legend.text=element_markdown(), legend.position = "inside",
        legend.justification.inside = c(0.98, 0.95),
        legend.background = element_rect(color='black'),
        plot.title.position = "plot", 
        plot.caption.position =  "plot") +
  ggtitle("A) Scan extraction with and without pre-loading")

pyopenms_precomp_gp <- read_csv("manuscript/data/pyopenms_precomp.csv") %>%
  pivot_longer(rep1:rep3) %>%
  mutate(type=factor(type, levels=c("init", "precomp"),
                       labels=fun_levels)) %>%
  mutate(method=factor(method, levels=c("spec", "chrom", "chrom_2D", "rtrange",
                                        "rtrange_2D", "premz", "fragmz"),
                       labels = c("Spectrum", "EIC", "EIC\n(2D Peak method)",
                                  "RT Range", "RT Range\n(2D Peak method)",
                                  "Precursor\nsearch", "Fragment\nsearch"))) %>%
  ggplot() +
  geom_boxplot(aes(x=method, y=value, color=type)) +
  scale_y_log10() +
  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
  theme_bw() +
  theme(legend.position = "none", 
        plot.title.position = "plot", 
        plot.caption.position =  "plot") +
  labs(x=NULL, y="Time (seconds)", color=NULL) +
  ggtitle("B) pyopenms metric timings with and without pre-loading")

init_outside_gp / pyopenms_precomp_gp
ggsave("manuscript/figures/supp_preload_fig.png", width = 6.5, height=5)
```

@@

We also investigated the efficacy of the various MS data formats for MS/MS data and found that support for fragmentation data searches was lacking or absent from the documentation and exposed functionality of each of these file types, requiring custom implementations every time. Despite both precursor searches (where all the precursors of a given fragment are found) and fragment searches (where all the fragments of a given precursor are found) representing intuitive and useful methods of MS/MS data processing, these timings were consistently among the slowest of the six query types for the non-database methods (Figure 2, rightmost panels).

All existing MS data types required multiple seconds to perform a single fragment search (Figure 2, bottom-right panel), representing a significant bottleneck for any downstream analysis requiring the data associated with the fragments of a given precursor. The SQL-based parsers, on the other hand, all took fractions of a second and consistently returned the relevant data hundreds of times more quickly than existing methods. The same was true for a precursor search across all methods aside from mzDB (Figure 2, bottom middle panel), which benefited significantly from constructing a single bounding box for all MS/MS information that requires a single decoding into computer memory, though this strategy will fail for any file with sufficiently large MS/MS data.

@@

File size is another important constraint on the efficacy of various MS formats. We measured the size on disk of each of the file types and found that they varied by approximately an order of magnitude, with HDF-based file types hovering around one-third the size of the mzML (83 megabytes, MB) while mzDB and mzMD were larger (108 MB and 96 MB, respectively). The SQLite object was the largest on disk of all the file types, nearly doubling the mzML's size at 145 MB, while DuckDB was on par with MZA at two-thirds of the mzML (54 MB) and Parquet improved slightly upon that again with its columnar-based storage format (Figure 2).

@@

The single-file, single-metric case discussed above and shown in Figure 2 is largely a worst-case scenario for many MS data systems that have a slow initial setup step to make downstream analysis faster. To compare these systems more fairly to our database schema, we also tested timings across multiple chromatograms. In each case, this was implemented as a for loop iterating over an increasing number of chromatograms corresponding to the largest intensity ions in the file (Figure 3).

```{r}
multichrom_times <- read_csv("manuscript/data/multichrom_times.csv")
timelabs <- c("0.01 seconds", "One second", "One minute", "One hour")
multichrom_times %>%
  mutate(method=factor(method, levels=method_names)) %>%
  ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
  geom_abline(slope=1, intercept=-3:3, color="black") +
  geom_smooth(method="lm", formula="y~x", alpha=0.2, fill="grey50") +
  geom_point(position = position_dodge(width=0.1), color="black", pch=21, size=3) +
  scale_y_log10(breaks=c(0.01, 1, 60, 3600), labels=timelabs) +
  scale_x_log10(breaks=c(1, 3, 10, 30, 100)) +
  scale_fill_manual(breaks = method_names, values=method_colors, 
                    aesthetics = c("fill", "color")) +
  labs(x="Number of chromatograms extracted", y="Time required",
       color=NULL, fill=NULL) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
ggsave("manuscript/figures/multichrom_fig.png", width = 6.5, height=5)
```

@@

*Figure 3: Scatter plot of the time required to extract multiple chromatograms using various methods on logarithmic axes. Best-fit linear models have been added for each method are shown behind triplicate timing measurements. Transparent intervals around each best-fit line show a single standard error of the mean. Chromatograms correspond to the largest intensity ions in the file. 1:1 lines have been added in black behind the data for comparison.*

Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure 3). pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30. Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure XX).
